Interactive Visualization of Hierarchically Structured Data

### 2016-12-06 treelapse.Rmd
Add more motivation in introduction

(todo: sentence about why we should care, and a high level summary
before diving into degree-of-interest linking jargon)

In this work, we describe methods for visualization of data structured
along trees, especially hierarchically structured collections of time
series. Influenced by Degree-of-Interest trees [@heer2004doitrees],
linked scatterplot brushing [@becker1987brushing], and timeboxes
[@hochheiser2004dynamic], we have adapted the fundamental
visualization principles of focus-plus-context and linking to study
questions that arise when analyzing tree structured data. Our
motivating application is to the analysis of microbial time series,
where an evolutionary tree relating microbes is available a
priori. However, we have identified common problem types where, if a
tree is not directly available, it can be constructed from data and
then studied using our techniques.

We have implemented our visualizations in d3, but encapsulated in an R
package, called treelapse, to facilitate rapid turnover from data
preparation and modeling to interactive exploration, and vice
versa. We hope this package encourages data analysts to work at the
border between data modeling and visualization, and more generally
empowers a wider audience to apply less widely known, but powerful,
visualization ideas.

@@

Finally, unrelated to established visualization principles, we note
that our work is deliberately grounded in the R software
ecosystem. Not only does this make these methods more broadly
accessible, we hope to facilitate exchange between data modeling and
interactive visualization. Moreover, our tools are intentionally
limited in scope -- designed to facilitate this dialog for a specific
class of problems, rather than providing a toolbox for generic types
of visualization design. We believe that this narrow context within a
broad ecosystem strikes a balance between problem-specificity and
ease-of-use.

@@

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow[^[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)]],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, computed monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See the Section
\@ref(conclusion) for potential optimizations, however.].

@@

```{r zillowmiddlepre}
knitr::include_graphics("zillow_middle_pre.png")
```

```{r zillowmiddleup}
knitr::include_graphics("zillow_middle_up.png")
```

```{r zillowmiddledown}
knitr::include_graphics("zillow_middle_down.png")
```

```{r zillowsf}
knitr::include_graphics("zillow_sf.png")
```

@@

Our next example is a study in bikesharing demand, included as an
example of analyzing collections of time series when there is no
obvious hierarchical structure a priori. The data are available at
the [UCI
Repository][^[(https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)]
and were originally collected by a Washington D.C.-based bikesharing
system for use in a Kaggle prediction competition. The data are hourly
measurements of bike demand, aggregated across all bikesharing
stations, over two years, along with supplemental weather data. In the
competition, participants were asked to predict the hourly demand on a
held-out test set. Here, we adopt a descriptive view instead,
attempting to characterize factors associated with variation in
bikesharing demand.

@@

Finally, we can study the behavior of the regression tree itself using
the DOI sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, Figures \@ref(fig:weekday2011) and
\@ref(fig:weekday2012) focuse on 2011 vs. 2012 split among weekday
samples, showing that this split distinguishes between samples falling
in the second and third quintiles of 8AM demand.

@@

>>>>

We introduce methods for visualization of data structured along trees,
especially hierarchically structured collections of time series. We
hope both to characterize generically useful techniques for
interactively visualizing hierarchical data and to offer practical
tools for implementing such displays. To this end, we identify
questions that often emerge when working with hierarchical data and
provide an R package to simplify their investigation [@ripley2001r].

Our key contribution is the adaptation of the visualization principles
of focus-plus-context and linking to the study of tree-structured data
[@buja1996interactive, @becker1987brushing]. Our motivating
application is to the analysis of microbial time series, where an
evolutionary tree relating microbes is available a priori. However, we
have identified common problem types where, if a tree is not directly
available, it can be constructed from data and then studied using our
techniques.

We have implemented our visualizations in D3, but encapsulated in an R
package, called treelapse, to facilitate rapid turnover from data
preparation and modeling to interactive exploration, and vice
versa. We hope this package encourages data analysts to work at the
border between data modeling and visualization, and more generally
empowers a wider audience to apply less widely known, but powerful,
visualization ideas.

@@

Finally, unrelated to established visualization principles, we note
that our work is deliberately grounded in the R software
ecosystem. This connection is made using the htmlwidgets package
[@vaidyanathan2014htmlwidgets]. Not only does linking R with D3 make
these visualization methods more broadly accessible, we hope to
facilitate exchange between data modeling and interactive
visualization. Moreover, our tools are intentionally limited in scope
-- designed to facilitate this dialog for a specific class of
problems, rather than providing a toolbox for generic types of
visualization design. We believe that this narrow context within a
broad ecosystem strikes a balance between problem-specificity and
ease-of-use.

@@

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow^[[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, computed monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See the Section
\@ref(conclusion) for potential optimizations, however.].

@@

```{r zillowmiddlepre, fig.cap = "Neighborhoods with mid-range home prices before the recession are selected here. Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
knitr::include_graphics("zillow_middle_pre.png")
```

```{r zillowmiddleup, fig.cap = "Among those neighborhoods with mid-range prices before the recession, we have selected those that recovered more rapidly. These appear to be located mainly in Los Angeles and San Diego counties."}
knitr::include_graphics("zillow_middle_up.png")
```

```{r zillowmiddledown, fig.cap = "Rather than selecting mid-range series that recovered quickly, we can isolate those whose prices remained depressed after the recession. These seem mostly to be located in Central California and the East San Francisco Bay Area."}
knitr::include_graphics("zillow_middle_down.png")
```

```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
knitr::include_graphics("zillow_sf.png")
```

@@

Our next example is a study in bikesharing demand, included as an
example of analyzing collections of time series when there is no
obvious hierarchical structure a priori. The data are available at
the UCI
Repository^[[https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)]
and were originally collected by a Washington D.C.-based bikesharing
system for use in a Kaggle prediction competition. The data are hourly
measurements of bike demand, aggregated across all bikesharing
stations, over two years, along with supplemental weather data. In the
competition, participants were asked to predict the hourly demand on a
held-out test set. Here, we adopt a descriptive view instead,
attempting to characterize factors associated with variation in
bikesharing demand.

@@

Finally, we can study the behavior of the regression tree itself using
the DOI sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, Figures \@ref(fig:weekday2011) and
\@ref(fig:weekday2012) focus on 2011 vs. 2012 split among weekday
samples, showing that this split distinguishes between samples falling
in the second and third quintiles of 8AM demand.

@@

Potential sections to write

* describe extension from svg to canvas
* give example of parallel coordinates / hierarchical clustering?
* explain incorporation into / give example of shiny app?
